A Hexagon Tile Map Algorithm for Displaying Spatial Data

### 2022-06-02 kobakian-cook.tex
links now working and more tweakig of paper

Alternative mapping methods, like cartograms, allow increased
understanding of the spatial distribution of a variable across the
population, by fairly representing the population in each administrative
area \citep{TAAM}. This acknowledges that the number of residents can be
different and also recognises that each area, or person within it is
equally important.

This paper contains a discussion of Existing Mapping Practices. This is
followed by details of the Algorithm. The implementation section
discusses the algorithm in the \CRANpkg{sugarbag} package. Finally, the
use and benefits of animation are presented.

\hypertarget{existing-mapping-practices}{%
\subsection{Existing Mapping
Practices}\label{existing-mapping-practices}}

@@

Administrative areas are often used to aggregate census data and used to
understand demographics within communities of the Australian population.
Each SA2 \citep{abs2011} was designed to represent a community. This
collection of communities presents an opportunity for explicit map
transformations to improve communication of spatial distributions
\citep{CBATCC}. In Figure \ref{fig:tas_displays} the choropleth map can
be seen underneath each alternative map display to allow for comparisons
to be made.

There are several established alternative visualisation methods. Tile
maps, rectangular cartograms \citep{ORC} and Dorling cartograms
\citep{ACTUC} all use one simple shape to represent each geographic
unit. They all minimise the emphasis on the size or shape of the
geographic areas. These alternative map displays focus on the
relationship between neighbours, attempting to preserve connections, and
disregard the unique shapes of the administrative boundaries. Figure
\ref{fig:tas_displays} shows a collection of alternative map displays,
this includes a) a contiguous cartogram, b) a non-contiguous cartogram,
and c) a Dorling cartogram.

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/tas_displays-1} \caption[The three displays show alternative maps of the Australian state of Tasmania at SA2 level]{The three displays show alternative maps of the Australian state of Tasmania at SA2 level: (a) contiguous cartogram, (b) non-contiguous cartogram and (c) Dorling cartogram of Tasmania. The contiguous cartogram looks like the state has an hourglass figure, while the non-contiguous cartogram shrinks areas into invisibility. The Dorling expands the metropolitan regions.}\label{fig:tas_displays}
\end{figure}
\end{Schunk}

When communicating information that is relevant to the population, each
member of the population can be given equal representation by
transforming the map \citep{TVSSS}. The connected communities can be
recognised by allowing the boundaries to maintain connection in the
transformed display. The contiguous cartogram displayed in Figure
\ref{fig:tas_displays}a) draws attention to smaller geographic units
when they are rescaled according to the population \citep{DMAHP}. These
new shapes can now be coloured to represent a second variable. The
algorithm uses the geographic shape of the areas and iterates toward
sizing the areas to represent the population. This display can create
twisted and unfamiliar shapes from the geographic units as the
algorithms must satisfy the topology conditions, especially when there
are communities located geographically far from their neighbours
\citep{TVSSS}.

The non-contiguous cartogram in Figure \ref{fig:tas_displays}b) also
uses the population to rescale the geographic units. Unlike the
contiguous cartogram, the SA2 areas maintain their geographic shape, but
they may not retain the connection with their neighbours. The population
of the SA2 areas is used to scale the geographic units in the
non-contiguous cartogram \citep{NAC}. The amount of background space can
be meaningful in non-contiguous cartograms \citep{ECGC}. The size
disparity between rural areas and urban areas result in reasonable
display within the south eastern city of Hobart, but these units are not
visible in the context of the Tasmania state map. Depending on the
difference between the population and geographic land size, the amount
of white space can also prevent meaningful understanding of the
distribution \citep{TVSSS}.

The Dorling cartogram presents each geographic unit as a circle, the
size of the circle is scaled according to the population value of each
area \citep{ACTUC}. Figure \ref{fig:tas_displays}c) shows the Tasmania
SA2 areas as an individual circle located as close as possible to the
geographic centroid location. This map draws attention to the collection
of coastal cities in Tasmania that were not apparent in Figure
\ref{fig:tas_displays} a), or b). This display also highlights that the
approximation of equal population is not easy to fulfil, as there is
some disparity in the sizes of the circles.

It is also common to treat all geographic units as equal, so that
regardless of the population or other information, to generate equal
size spatial polygons. The Dorling algorithm can beinduced to do this,
to make equallly sized circles. \citet{MDAC} describes a group of
alternative maps called ``mosaic cartograms'' for tile displays. The
geography of the USA has been transformed into a square-binned tile map
manually and made available in the R package \CRANpkg{statebins}
\citep{statebins}.The algorithm described below is an automatic layout
of hexagon tiles, where a continguity constraint is relaxed. This gives
equal weight to each geographic area even if the population differs.

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/hexmap-1} \caption[A hexagon tile map of female tyroid cancer incidence in Australia, the same data as shown in the choropleth map in Figure 1]{A hexagon tile map of female tyroid cancer incidence in Australia, the same data as shown in the choropleth map in Figure 1. The high incidence in several of the metropolitan regions (Brisbane, Sydney and Perth) can now be seen, along with numerous isolated spots.}\label{fig:hexmap}
\end{figure}
\end{Schunk}

There are several key steps in the creation of the hexagon tile map as
described in the flow chart in Figure \ref{fig:sugarbag_flow}. First,
derive the set of centroids from the polygons provided, then create the
grid of hexagon locations. These two processes are defined in the blue
left column of the flow chart in figure \ref{fig:sugarbag_flow}. Each
centroid can then be allocated to an available hexagon location. The
steps for the allocation process are detailed in the right column of
Figure \ref{fig:sugarbag_flow}. There are several filter steps to speed
the process of selecting an appropriate hexagon to represent each
geographic unit. To make tessellated plots with the hexagon allocations,
the point locations are converted into hexagon shapes.

\begin{figure}
\includegraphics[width=14cm]{figs/sugarbag flow.png}
\caption{\label{fig:sugarbag_flow}A flow diagram detailing the steps taken to create a hexagon tile map. There are two basic processes, one to make the grid, and the other to allocate centroids to grid points.}
\end{figure}

@@

A grid is created to allow tessellation of the hexagons that represent
the geographic units. For a hexagon tile map, the grid of possible
hexagon locations is made using the \code{create\_grid} function. It
uses the centroids, the hexagon size and the buffer distance. Figure
\ref{fig:filterprocess} 1a) shows the initial locations of grid points
created for Tasmania.

@@

A set of longitude columns, and latitude rows are created to define the
locations of the hexagons. The distance between each row and column is
the size specified by \code{hex\_size}. Equally spaced columns are
created from the minimum longitude minus the buffer distance, up to the
maximum longitude plus the buffer distance. Similarly, the rows are
created from the latitude values and the buffer distance. A unique
hexagon location is created from all intersections of the longitude
columns and latitude rows. Figure \ref{fig:filterprocess} 1a) shows the
hexagon grid after every second latitude row on the grid is shifted
right, by half of the hexagon size.

@@

The hexagon grid points are kept only if they fall within the rolling
average of the minimum and maximum centroid values after accounting for
the buffer distance, for each row and column of the grid. Figure
\ref{fig:filterprocess} 1b) displays remaining hexagon grid points after
applying the buffer filter. The sparsely populated South-West region of
National Park has fewer points available in the water compared to the
South-East region near the city of Hobart.

@@

The distance is calculated between each centroid in the set, and each of
the focal points provided. The order for allocation is determined by the
distance between the polygon centroid and it's closest focal point. For
this example, this distance is only calculated for the capital city of
Hobart, represented in figure \ref{fig:filterprocess} 2a) and 2b) as a
white cross.

@@

@@

The set of possible hexagon grid points reduces by one after each
centroid is allocated, then only the grid points that have not yet been
selected are considered. Keeping only the available hexagon points
prevents multiple geographic units from being allocated to the same
hexagon. This is demonstrated in figure \ref{fig:filterprocess} 2a) and
2b) by the black hexagons that represent the seven closest polygons to
the capital city of Hobart. As the allocation process begins for the
eighth closest centroid there are seven unavailable hexagon locations.

@@

The \code{width} parameter is used to take a slice of the remaining
points. The width is the amount of degrees used on either side of the
angle from the focal point to centroid location. This uses the angle
from the closest capital city, to the current centroid as seen in figure
\ref{fig:filterprocess} 2a). This allows the spatial relationship to be
preserved, even when it is allocated to a hexagon that is further from
the focal point then the original centroid location.

@@

This algorithm should be general enough to apply to other study regions
and contexts. The algorithm has only been tested using single countries.
While the buffer allows extension beyond the furthest centroids, there
is no mechanism implemented to protect the borders and ensure centroids
are placed within the geographic borders of a country. The algorithm is
also currently limited to representing each geographic area with only
one hexagon. Future developments may address these limitations, and
refine the filter steps for final selection of a hexagon grid point for
each centroid. A direct angle is currently implemented, but may narrowly
miss appropriate hexagons, a logarithmic function may help to choose a
closer hexagon to the original centroid location. This algorithm is an
effective start to automatically creating hexagon tile maps for many
geographic units. This contributes to an extensive body of work that
encourages the use of alternative map displays for statistical displays
of spatial data.

>>>>

Alternative mapping methods, like cartograms implemented in
\CRANpkg{cartogram}, allow increased understanding of the spatial
distribution of a variable across the population, by fairly representing
the population in each administrative area \citep{TAAM}. This
acknowledges that the number of residents can be different and also
recognises that each area, or person within it is equally important.

This paper contains a discussion of existing mapping practices, followed
by details of the algorithm. The implementation section explains the
algorithm available in the \CRANpkg{sugarbag} package. How to utilise
animation with the hexagon map is described. The paper finishes with a
summary and possible new directions for the algorithm.

\hypertarget{existing-mapping-practices}{%
\subsection{Existing mapping
practices}\label{existing-mapping-practices}}

@@

Administrative areas are often used to aggregate census data and used to
understand demographics within communities of the Australian population.
Each SA2 \citep{abs2011} was designed to represent a community. This
collection of communities presents an opportunity for explicit map
transformations to improve communication of spatial distributions
\citep{CBATCC}. In Figure \ref{fig:tasdisplays} the choropleth map can
be seen underneath each alternative map display to allow for comparisons
to be made.

There are several established alternative visualisation methods.
Rectangular cartograms \citep{ORC} and Dorling cartograms \citep{ACTUC}
implemented in \pkg{cartogram} and tile maps implemented in
\CRANpkg{tilemaps}, all use one simple shape to represent each
geographic unit. They all minimise the emphasis on the size or shape of
the geographic areas. These alternative map displays focus on the
relationship between neighbours, attempting to preserve connections, and
disregard the unique shapes of the administrative boundaries. Figure
\ref{fig:tasdisplays} shows a collection of alternative map displays,
this includes a) a contiguous cartogram, b) a non-contiguous cartogram,
and c) a Dorling cartogram.

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{kobakian-cook_files/figure-latex/tasdisplays-1} \caption[The three displays show alternative maps of the Australian state of Tasmania at SA2 level]{The three displays show alternative maps of the Australian state of Tasmania at SA2 level: (a) contiguous cartogram, (b) non-contiguous cartogram and (c) Dorling cartogram of Tasmania. The contiguous cartogram looks like the state has an hourglass figure, while the non-contiguous cartogram shrinks areas into invisibility. The Dorling expands the metropolitan regions.}\label{fig:tasdisplays}
\end{figure}
\end{Schunk}

When communicating information that is relevant to the population, each
member of the population can be given equal representation by
transforming the map \citep{TVSSS}. The connected communities can be
recognised by allowing the boundaries to maintain connection in the
transformed display. The contiguous cartogram displayed in Figure
\ref{fig:tasdisplays}a draws attention to smaller geographic units when
they are rescaled according to the population \citep{DMAHP}. These new
shapes can now be coloured to represent a second variable. The algorithm
uses the geographic shape of the areas and iterates toward sizing the
areas to represent the population. This display can create twisted and
unfamiliar shapes from the geographic units as the algorithms must
satisfy the topology conditions, especially when there are communities
located geographically far from their neighbours \citep{TVSSS}.

The non-contiguous cartogram in Figure \ref{fig:tasdisplays}b also uses
the population to rescale the geographic units. Unlike the contiguous
cartogram, the SA2 areas maintain their geographic shape, but they may
not retain the connection with their neighbours. The population of the
SA2 areas is used to scale the geographic units in the non-contiguous
cartogram \citep{NAC}. The amount of background space can be meaningful
in non-contiguous cartograms \citep{ECGC}. The size disparity between
rural areas and urban areas result in reasonable display within the
south eastern city of Hobart, but these units are not visible in the
context of the Tasmania state map. Depending on the difference between
the population and geographic land size, the amount of white space can
also prevent meaningful understanding of the distribution \citep{TVSSS}.

The Dorling cartogram presents each geographic unit as a circle, the
size of the circle is scaled according to the population value of each
area \citep{ACTUC}. Figure \ref{fig:tasdisplays}c shows the Tasmania SA2
areas as an individual circle located as close as possible to the
geographic centroid location. This map draws attention to the collection
of coastal cities in Tasmania that were not apparent in Figure
\ref{fig:tasdisplays} a or b. This display also highlights that the
approximation of equal population is not easy to fulfil, as there is
some disparity in the sizes of the circles.

It is also common to treat all geographic units as equal, so that
regardless of the population or other information, to generate equal
size spatial polygons. The Dorling algorithm can be induced to do this,
to make equallly sized circles. \citet{MDAC} describes a group of
alternative maps called ``mosaic cartograms'' for tile displays. The
geography of the USA has been transformed into a square-binned tile map
manually and made available in the R package \CRANpkg{statebins}
\citep{statebins}.The algorithm described below is an automatic layout
of hexagon tiles, where a continguity constraint is relaxed. This gives
equal weight to each geographic area even if the population differs.

The nature of Australian population boundaries means that there are many
connecting neighbours of population areas. This solution was designed to
visualise Australian population analysis, inspired by tile maps three
sided triangles and four sided squares were considered, but this would
not have allowed sufficient neighbours to connect. Hexagon tile maps
utilise the tessellating features of hexagons to maximise the data ink
and allow connectivity between six potential neighbours. Hexagons allow
the alignment of geographic borders, this aides in comparison of colour
values.

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/hexmap-1} \caption[A hexagon tile map of female thyroid cancer incidence in Australia, the same data as shown in the choropleth map in Figure 1]{A hexagon tile map of female thyroid cancer incidence in Australia, the same data as shown in the choropleth map in Figure 1. The high incidence in several of the metropolitan regions (Brisbane, Sydney and Perth) can now be seen, along with numerous isolated spots.}\label{fig:hexmap}
\end{figure}
\end{Schunk}

There are several key steps in the creation of the hexagon tile map as
described in the flow chart in Figure \ref{fig:sugarbagflow}. First,
derive the set of centroids from the polygons provided, then create the
grid of hexagon locations. These two processes are defined in the blue
left column of the flow chart in figure \ref{fig:sugarbagflow}. Each
centroid can then be allocated to an available hexagon location. The
steps for the allocation process are detailed in the right column of
Figure \ref{fig:sugarbagflow}. There are several filter steps to speed
the process of selecting an appropriate hexagon to represent each
geographic unit. To make tessellated plots with the hexagon allocations,
the point locations are converted into hexagon shapes.

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{figs/sugarbag_flow} \caption[A flow diagram detailing the steps taken to create a hexagon tile map]{A flow diagram detailing the steps taken to create a hexagon tile map. There are two basic processes, one to make the grid, and the other to allocate centroids to grid points.}\label{fig:sugarbagflow}
\end{figure}
\end{Schunk}

@@

A grid is created to allow tessellation of the hexagons that represent
the geographic units. For a hexagon tile map, the grid of possible
hexagon locations is made using the \code{create\_grid} function. It
uses the centroids, the hexagon size and the buffer distance. Figure
\ref{fig:filterprocess} 1a shows the initial locations of grid points
created for Tasmania.

\begin{verbatim}
centroids <- create_centroids(
  shp_sf = sa2 %>% filter(state_name_2011 == "Tasmania"), 
  sf_id = "sa2_name_2011")
\end{verbatim}

@@

A set of longitude columns, and latitude rows are created to define the
locations of the hexagons. The distance between each row and column is
the size specified by \code{hex\_size}. Equally spaced columns are
created from the minimum longitude minus the buffer distance, up to the
maximum longitude plus the buffer distance. Similarly, the rows are
created from the latitude values and the buffer distance. A unique
hexagon location is created from all intersections of the longitude
columns and latitude rows. Figure \ref{fig:filterprocess} 1a shows the
hexagon grid after every second latitude row on the grid is shifted
right, by half of the hexagon size.

\begin{verbatim}
grid <- create_grid(centroids = centroids, hex_size = 0.2, buffer_dist = 2)
\end{verbatim}

@@

The hexagon grid points are kept only if they fall within the rolling
average of the minimum and maximum centroid values after accounting for
the buffer distance, for each row and column of the grid. Figure
\ref{fig:filterprocess} 1b displays remaining hexagon grid points after
applying the buffer filter. The sparsely populated South-West region of
National Park has fewer points available in the water compared to the
South-East region near the city of Hobart.

@@

The distance is calculated between each centroid in the set, and each of
the focal points provided. The order for allocation is determined by the
distance between the polygon centroid and it's closest focal point. For
this example, this distance is only calculated for the capital city of
Hobart, represented in Figure \ref{fig:filterprocess} 2a and 2b as a
white cross.

@@

\begin{verbatim}
hex_allocated <- allocate(centroids = centroids,
                          sf_id = "SA2_NAME16",
                          hex_grid = grid,
                          hex_size = 0.2, # same size used in create_grid
                          hex_filter = 10,
                          use_neighbours = tas_sa2,
                          focal_points = capital_cities,
                          # same column used in create_centroids
                          width = 30, verbose = TRUE)
\end{verbatim}

@@

The set of possible hexagon grid points reduces by one after each
centroid is allocated, then only the grid points that have not yet been
selected are considered. Keeping only the available hexagon points
prevents multiple geographic units from being allocated to the same
hexagon. This is demonstrated in Figure \ref{fig:filterprocess} 2a and
2b by the black hexagons that represent the seven closest polygons to
the capital city of Hobart. As the allocation process begins for the
eighth closest centroid there are seven unavailable hexagon locations.

@@

The \code{width} parameter is used to take a slice of the remaining
points. The width is the amount of degrees used on either side of the
angle from the focal point to centroid location. This uses the angle
from the closest capital city, to the current centroid as seen in Figure
\ref{fig:filterprocess} 2a. This allows the spatial relationship to be
preserved, even when it is allocated to a hexagon that is further from
the focal point then the original centroid location.

@@

Australia is used as an example, because it represents an extreme sparse
geography and small dense population areas which produces unsatisfactory
results with all existing cartogram-type algorithms. However, the
algorithm is sufficiently general to apply to other geographic regions,
and indeed has been applied to make hexagon maps of the UK, USA and
France.

There are many possible extensions to the algorithm that might be
valuable future work. The buffer allows extension beyond the furthest
centroids, but there is no mechanism to protect borders to force
centroids are placed within the geographic borders of a region
(e.g.~country or state). The algorithm is also currently limited to
representing each geographic area with only one hexagon. In the filter
step for final selection of a hexagon grid point for each centroid, a
direct angle constraint is currently implemented, but it may narrowly
miss potentially better hexagon allocation. Some other constraints, such
as a logarithmic function may help to choose refine the final allocation
to hexagons closer to the original centroid location. In a country like
Australia, the vast centre produces a hexagon map with small isolated
hexagons, even when the city centres are hugely expanded. It could be
useful to allow variable sized hexagons, primarily to increase the size
of the isolated hexagons to improve their visibility.

This new algorithm provides an effective start point for automatically
creating hexagon tile maps for any spatial polygon data, and contributes
to an extensive body of work that encourages the use of alternative map
displays for statistical displays of spatial data.
